Computer-Assisted Analysis of the \\ Anderson-Hájek Ontological Controversy

### 2017-01-02 Papers/2017/LogicaUniversalis-HajekAndersonControversy/main.tex
modifications requested by report 1

\noindent The axioms in G\"odel's ontological proof
\citep{GoedelNotes,ScottNotes} (cf. Fig. \ref{fig:Goedel}) entail what is called \emph{modal collapse}
\citep{Sobel1987,SobelBook2004}: the formula $\varphi \rightarrow \Box
\varphi$, abbreviated as MC, holds for any formula $\varphi$ and not
just for $\exists x. \mathit{God}(x)$ as intended.

\begin{figure}[t]
\noindent \framebox[\columnwidth][r]{
\begin{minipage}{.94\columnwidth}\small
\begin{itemize}
\item[A1] Either a property or its negation is positive, but not
  both:
  $$\hol{\allq \varphi [P(\neg \varphi) \biimp \neg P(\varphi)]}$$
\item[A2] A property necessarily implied by a
  positive property is positive:
  $$\hol{\allq \varphi \allq \psi [(P(\varphi) \wedge \nec \allq x [\varphi(x)
  \imp \psi(x)]) \imp P(\psi)]}$$
\item[T1] Positive properties are possibly exemplified:
  $$\hol{\allq \varphi [P(\varphi) \imp \pos \exq x \varphi(x)]}$$
\item[D1] A \emph{God-like} being possesses all positive properties:
  $$\hol{G(x) \equiv \forall \varphi [P(\varphi) \imp \varphi(x)]}$$
\item[A3]  The property of being God-like is positive:
  $$\hol{P(G)}$$
\item[C\phantom{1}] Possibly, a God-like being exists: $$\hol{\pos \exq x G(x)}$$
\item[A4]  Positive properties are necessarily positive:
  $$\hol{\allq \varphi [P(\varphi) \imp \Box \; P(\varphi)]}$$
\item[D2] An \emph{essence} of an individual is a property possessed by it and necessarily implying any of its properties: $$\hol{\ess{\varphi}{x} \equiv \varphi(x) \wedge \allq
  \psi (\psi(x) \imp \nec \allq y (\varphi(y) \imp \psi(y)))}$$
\item[T2]  Being God-like is an essence of any
  God-like being: $$\hol{\allq x [G(x) \imp \ess{G}{x}]}$$
\item[D3] \emph{Necessary existence} of an individual is the necessary exemplification of all its essences:
  $$\hol{\NE(x) \equiv \allq \varphi [\ess{\varphi}{x} \imp \nec
  \exq y \varphi(y)]}$$
\item[A5] Necessary existence is a positive property: $$\hol{P(\NE)}$$
\item[L1] If a God-like being exists, then necessarily a God-like being exists:
  $$\hol{\exq x G(x) \imp \nec \exq y G(y)}$$
\item[L2] If possibly a God-like being exists, then necessarily a God-like being exists:
  $$\hol{\pos \exq x G(x) \imp \nec \exq y G(y)} $$
%
\item[T3] Necessarily, a God-like being exists: $$\hol{\nec \exq x G(x)}$$
\end{itemize}
\end{minipage}
} \vskip-.5em
\caption{Scott's version of G\"odel's ontological argument\label{fig:Goedel}}
\end{figure}

This fact, which
has recently been confirmed with higher-order automated theorem
provers \citep{C40,J30}, has led to strong criticism of the argument
and stimulated attempts to remedy the problem.
\citet{Hajek_Magari_and_others_1996,Hajek_der_Mathematiker_2001}
proposed the use of cautious instead of full comprehension principles,
and \citet{fitting02:_types_tableaus_god} took
greater care of the semantics of higher-order quantifiers
in the presence of modalities. Others, such as
\citet{anderson90:_some_emend_of_goedel_ontol_proof}, \citet{Hajek2002}
and \citet{bjordal99}, proposed emendations of G\"odel's
axioms and definitions. They require neither comprehension
restrictions nor more complex semantics. Therefore, they are
technically simpler to analyze with computer support. We have
formalized those using the proof assistant Isabelle/HOL \citep{Isabelle}
together with the automated higher-order reasoners Leo-II \citep{C26},
Satallax \citep{brown2012satallax}, Metis
\citep{Hurd03first-orderproof}, and Nitpick \citep{Nitpick}.   Our
formalizations\footnote{The   formalizations are available in the
subdirectories \url{Anderson}, \url{Hajek} and \url{Bjordal} at \url{gi
thub.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/
}.} employ the embedding of higher-order modal logic (HOML) in
classical higher-order logic (HOL) as introduced in previous work
\citep{C40,J30,J23}. We explored the effect of different domain
conditions on the provability of lemmas, theorems and even axioms.
This was motivated by a controversy between Hájek and Anderson
regarding the redundancy of some axioms in Anderson's emendation. In
\emph{constant domain semantics}, the individual domains are the same
in all possible worlds. In \emph{varying domain semantics}, the
domains may vary from world to world. This variation is technically
encoded with the help of an existence relation expressing which
individuals actually exist in each world. Quantifiers are then
uniformly formalized as \emph{actualistic quantifiers} (i.e. guarded by
the existence relation). Our main results are summarized here.

For all emendations and variants discussed here, the axioms and
definitions have been shown to be consistent and not to entail modal
collapse.

@@

Using our approach, the formalization and (partly) automated analysis
of several variants of G\"odel's ontological argument has been
surprisingly straightforward. The provers not only confirmed many
claimed results, but also exposed a few mistakes and novel insights
in a long standing controversy.
We believe the technology employed in
this work is ready to be fruitfully adopted in larger scale by
philosophers.

>>>>

\section{Introduction}
We present an exemplary study in \emph{Computational Metaphysics},
which is an emerging, interdisciplinary field aiming at the rigorous
formalisation and deep logical assessment of rational, philosophical
arguments on the computer.  The particular focus here is on the
ontological argument for the existence of God. More precisely, we
further expand our studies on modern (modal) variants of the
ontological argument that are essentially emendations of Kurt Gödel's
seminal contribution \citep{GoedelNotes}, respectively, Dana Scott's
variant of it \cite{ScottNotes}. In related experiments
\citep{C55,C40,J30}, we have already demonstrated that our technology
is suited not only to formally assess the logical quality of the
presented arguments, but it may even generate new knowledge about
it. For example, the theorem prover Leo-II \cite{Leo-II} detected an
inconsistency in Gödel's original variant \citep{GoedelNotes} of the
argument, which is not the case for Scott \citep{ScottNotes}.  
We here assume some familiarity with these related experiments.

The particular motivation and context for work presented in this
article is as follows: The axioms in G\"odel's ontological proof
\citep{GoedelNotes,ScottNotes} (cf. Fig. \ref{fig:Goedel}) entail what
is called \emph{modal collapse} \citep{Sobel1987,SobelBook2004}: the
formula $\varphi \rightarrow \Box \varphi$, abbreviated as MC, holds
for any formula $\varphi$ and not just for
$\exists x. \mathit{God}(x)$ as intended.
\begin{figure}[t]
\noindent \framebox[\columnwidth][r]{
\begin{minipage}{.94\columnwidth}\small
\begin{itemize}
\item[A1] Either a property or its negation is positive, but not
  both:
  $$\hol{\allq \varphi [P(\neg \varphi) \biimp \neg P(\varphi)]}$$
\item[A2] A property necessarily implied by a
  positive property is positive:
  $$\hol{\allq \varphi \allq \psi [(P(\varphi) \wedge \nec \allq x [\varphi(x)
  \imp \psi(x)]) \imp P(\psi)]}$$
\item[T1] Positive properties are possibly exemplified:
  $$\hol{\allq \varphi [P(\varphi) \imp \pos \exq x \varphi(x)]}$$
\item[D1] A \emph{God-like} being possesses all positive properties:
  $$\hol{G(x) \equiv \forall \varphi [P(\varphi) \imp \varphi(x)]}$$
\item[A3]  The property of being God-like is positive:
  $$\hol{P(G)}$$
\item[C\phantom{1}] Possibly, a God-like being exists: $$\hol{\pos \exq x G(x)}$$
\item[A4]  Positive properties are necessarily positive:
  $$\hol{\allq \varphi [P(\varphi) \imp \Box \; P(\varphi)]}$$
\item[D2] An \emph{essence} of an individual is a property possessed by it and necessarily implying any of its properties: $$\hol{\ess{\varphi}{x} \equiv \varphi(x) \wedge \allq
  \psi (\psi(x) \imp \nec \allq y (\varphi(y) \imp \psi(y)))}$$
\item[T2]  Being God-like is an essence of any
  God-like being: $$\hol{\allq x [G(x) \imp \ess{G}{x}]}$$
\item[D3] \emph{Necessary existence} of an individual is the necessary exemplification of all its essences:
  $$\hol{\NE(x) \equiv \allq \varphi [\ess{\varphi}{x} \imp \nec
  \exq y \varphi(y)]}$$
\item[A5] Necessary existence is a positive property: $$\hol{P(\NE)}$$
\item[L1] If a God-like being exists, then necessarily a God-like being exists:
  $$\hol{\exq x G(x) \imp \nec \exq y G(y)}$$
\item[L2] If possibly a God-like being exists, then necessarily a God-like being exists:
  $$\hol{\pos \exq x G(x) \imp \nec \exq y G(y)} $$
%
\item[T3] Necessarily, a God-like being exists: $$\hol{\nec \exq x G(x)}$$
\end{itemize}
\end{minipage}
} \vskip-.5em
\caption{Scott's version of G\"odel's ontological argument\label{fig:Goedel}}
\end{figure}
This fact, which has been confirmed with higher-order automated
theorem provers \citep{C40,J30} in our previous experiments, has led
to strong criticism of the argument and stimulated attempts to remedy
the problem.
\citet{Hajek_Magari_and_others_1996,Hajek_der_Mathematiker_2001}
proposed the use of cautious instead of full comprehension principles,
and \citet{fitting02:_types_tableaus_god} took greater care of the
semantics of higher-order quantifiers in the presence of
modalities. Others, such as
\citet{anderson90:_some_emend_of_goedel_ontol_proof},
\citet{Hajek2002} and \citet{bjordal99}, proposed emendations of
G\"odel's axioms and definitions. They require neither comprehension
restrictions nor more complex semantics. Therefore, they are
technically simpler to analyze with computer support. We have
formalized those using the proof assistant Isabelle/HOL
\citep{Isabelle} together with the automated higher-order reasoners
Leo-II \citep{Leo-II}, Satallax \citep{brown2012satallax}, Metis
\citep{Hurd03first-orderproof}, and Nitpick \citep{Nitpick}.  Our
formalizations employ the embedding of higher-order modal logic (HOML)
in classical higher-order logic (HOL) as introduced in previous work
\citep{C40,J23}. We explored the effect of different domain conditions
on the provability of lemmas, theorems and even axioms.  This was
motivated by a controversy between Hájek and Anderson regarding the
redundancy of some axioms in Anderson's emendation. In \emph{constant
  domain semantics}, the individual domains are the same in all
possible worlds. In \emph{varying domain semantics}, the domains may
vary from world to world. This variation is technically encoded with
the help of an existence relation expressing which individuals
actually exist in each world. Quantifiers are then uniformly
formalized as \emph{actualistic quantifiers} (i.e. guarded by the
existence relation). \footnote{EDNOTE: This is misleading statement;
  we need to reword}
Our main results, which we have presented at the
1st World Congress on Logic and Religion \citep{C41}, are summarized
below. The corresponding formalizations, containing all the details,
are available in the subdirectories \url{Anderson}, \url{Hajek} and
\url{Bjordal} at
\url{github.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/}.

\section{Results from our Experiments}
For all emendations and variants discussed here, the axioms and
definitions have been shown to be consistent and not to entail modal
collapse.

@@

\section{Conclusion}
Using our approach (semantical embeddings of variants of higher-order
modal logics in classical higher-order logic), the formalization and
(partly) automated analysis of several variants of G\"odel's
ontological argument has been surprisingly straightforward. The
provers not only confirmed many claimed results, but also exposed a
few mistakes and novel insights in a long standing controversy.  We
believe the technology employed in this work is ready to be fruitfully
adopted in larger scale by philosophers. Moreover, the technology is
by now means restricted to metaphysics or philosophy, and further work
includes similar applications in other areas, including, for example,
law, politics and ethics.
